---
title: "Using table.express"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using table.express}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library("table.express")
library("data.table")
library("magrittr")

# knitr defaults
knitr::opts_chunk$set(comment = "#>")

options("datatable.print.nrows" = 4L)
options("datatable.print.topn" = 2L)
```

The goal of this package is to offer an alternative way of **expressing** common operations with
[`data.table`](https://cran.r-project.org/package=data.table)
without sacrificing the performance optimizations that it offers.
The foundation for the data manipulation verbs is the
[`dplyr`](https://cran.r-project.org/package=dplyr)
package,
which also advocates the piping operator from the
[`magrittr`](https://cran.r-project.org/package=magrittr)
package.
The [`rlang`](https://cran.r-project.org/package=rlang)
package powers most of this package's functionality,
which means that
[tidy evaluation](https://tidyeval.tidyverse.org/)
can also be supported.
There are other resources that provide comprehensive descriptions of these packages,
so they will not be explained here.

Even though using data manipulation verbs can improve expressiveness in some cases,
this is not always true,
so using the traditional `data.table` syntax might still be preferable in many situations.

In order to resemble SQL syntax more closely,
a couple of verb aliases are also defined:

- `where`, being equivalent to `dplyr`'s `filter`.
- `order_by`, being equivalent to `dplyr`'s `arrange`.

The examples here will be working with the `mtcars` data:

```{r mtcars}
data("mtcars")

DT <- mtcars %>%
  as.data.table %T>%
  print
```

## Expression delimiters

The foundation for this package is building expressions that are almost entirely delegated to `data.table`.
These expressions are built by *parsing* the input of the different verbs.
In order to explicitly show when an expression is being built and subsequently evaluated,
we use 3 delimiters:

- `start_expr`
- `end_expr`
- `chain`

These also serve as visual reminders that we are not dealing directly with `data.table`s during the process.
We capture the input `data.table` and start the process with `start_expr`,
and evaluate the final expression with `end_expr`.
Using `chain` is equivalent to calling `end_expr` immediately followed by `start_expr`.

## Arranging rows

The `arrange`/`order_by` verbs add an expression with `order` to the frame,
and let `data.table` handle it as usual:

```{r arrange}
DT %>%
  start_expr %>%
  order_by(mpg, -cyl) %T>%
  print %>%
  end_expr
```

We see here that the built expression includes a `.DT_` pronoun.
When the expression is evaluated,
the captured `data.table` is assigned to the evaluation environment as said pronoun.

## Selecting columns

Even though selecting a subset of columns is a common operation,
it may be undesirable to do so when working with `data.table`s because they lead to data copies.
Given that `data.table` supports modification by reference,
these copies are necessary.

With that said,
the `select` verb can be used as usual:

```{r select}
DT %>%
  start_expr %>%
  select(mpg, am) %T>%
  print %>%
  end_expr
```

To maintain consistency,
even single columns are kept as `data.table`s:

```{r select-single}
DT %>%
  start_expr %>%
  select(mpg) %T>%
  print %>%
  end_expr
```

In the case of single expressions in `select`,
calls to `tidyselect`'s helpers or to `:` are handled specially internally:

```{r select-special}
DT %>%
  start_expr %>%
  select(mpg:cyl) %>%
  end_expr

DT %>%
  start_expr %>%
  select(contains("M", ignore.case = TRUE)) %>%
  end_expr
```

Tidy evaluation and the `.parse` argument can also aid in cases where certain parts of the frame were computed programmatically:

```{r select-tidy}
selected <- c("mpg", "am")
DT %>%
  start_expr %>%
  select(!!!selected, .parse = TRUE) %>%
  end_expr
```

## Transmuting columns

Given the way `data.table` handles the `j` part of the frame,
creating and keeping only new columns
(like with `dplyr`'s `transmute`)
can be done with `select`,
so `transmute` is simply an alias during expression building.

```{r transmute}
DT %>%
  start_expr %>%
  select(foo = mpg * 2, bar = exp(cyl)) %>%
  end_expr

DT %>%
  start_expr %>%
  transmute(foo = mpg * 2, bar = exp(cyl)) %>%
  end_expr
```

## Mutating columns

The `mutate` verb builds an expression with `:=` in order to perform assignment by reference by default.
This can be avoided by passing `.by_ref = FALSE` to `end_expr`,
which will use `data.table::copy` before assigning `.DT_`:

```{r mutate}
DT %>%
  start_expr %>%
  mutate(mpg = mpg / 2, hp = log(hp))

DT %>%
  start_expr %>%
  mutate(mpg = mpg / 2, hp = log(hp)) %>%
  end_expr(.by_ref = FALSE) %>% {
    invisible(print(.))
  }

print(DT)
```

## Filtering rows

The `where`/`filter` verbs work with the `i` part of the frame:

```{r filter}
DT %>%
  start_expr %>%
  filter(vs == 1L, carb > 2L) %T>%
  print %>%
  end_expr

DT %>%
  start_expr %>%
  select(mean_mpg = mean(mpg)) %>%
  where(vs == 1L, carb > 2L, .collapse = `|`) %T>%
  print %>%
  end_expr
```

The helper verb `filter_sd` can be used to apply the same conditions to many columns,
and it can use a special pronoun `.COL` while specifying the expression,
as well as `tidyselect` helpers to choose `.SDcols`
(with caveats, see [eager verbs](#eager-verbs)):

```{r filter-sd}
DT %>%
  start_expr %>%
  filter_sd(`>`, 20, .SDcols = c("mpg", "qsec")) %T>%
  print %>%
  end_expr

DT %>%
  start_expr %>%
  filter_sd(.COL > 20, .SDcols = c("mpg", "qsec")) %>%
  end_expr

DT %>%
  start_expr %>%
  filter_sd(.COL > 0, .SDcols = contains("m"))
```

### Using keys or secondary indices

The `filter_on` verb can be used to build an expression that specifies the `on` argument of the frame.
It accepts key-value pairs where each key is a column in the data,
and each value is the corresponding value that the column should have to match:

```{r filter-on}
DT %>%
  start_expr %>%
  filter_on(cyl = 6, am = 0) %T>%
  print %>%
  end_expr

DT %>%
  start_expr %>%
  filter_on(cyl = 6, am = 0, mult = "first") %>%
  end_expr
```

## Modifying subset of data

In order to support functionality similar to `data.table`'s `DT[, lapply(.SD, fun), .SDcols = c("...")]` syntax,
2 `data.table`-specific verbs are provided: `mutate_sd` and `transmute_sd`.

Starting with `mutate_sd`,
it modifies columns in `.SDcols` by reference,
and columns that are not part of `.SDcols` are kept:

```{r mutate-sd}
DT %>%
  start_expr %>%
  mutate_sd(exp, .SDcols = c("mpg", "cyl")) %>%
  end_expr

print(DT)
```

Additionally, `mutate_sd` supports the special `.COL` pronoun that symbolizes the column that should be modified,
and can be used to express the mutation expression:

```{r mutate-sd-col}
DT %>%
  start_expr %>%
  mutate_sd(log(.COL), .SDcols = c("mpg", "cyl")) %>%
  end_expr

print(DT)
```

On the other hand, `transmute_sd` never modifies by reference,
and supports special expressions to "build" what is chosen as `.SDcols`.
These expressions can use `tidyselect` helpers,
as well as another special pronoun: `.COLNAME`:

```{r transmute-sd}
DT %>%
  start_expr %>%
  transmute_sd(.COL * 2, .SDcols = starts_with("d")) %>%
  end_expr

DT %>%
  start_expr %>%
  transmute_sd(.COL * 2, .SDcols = grepl("^d", .COLNAME)) %>%
  end_expr
```

## Data manipulation by group

Since `data.table` already supports this by means of its `by` parameter,
the `group_by` verb simply parses its input and assigns it accordingly:

```{r group-by}
DT %>%
  start_expr %>%
  select(.N) %>%
  group_by(gear) %T>%
  print %>%
  end_expr
```

The `key_by` verb does the same but sets the key of the result in order to sort:

```{r key-by}
DT %>%
  start_expr %>%
  select(.N) %>%
  key_by(gear) %T>%
  print %>%
  end_expr
```

## Automatic expression chaining

A `data.table`'s frame has 3 main elements:
[`i`, `j`, and `by`](https://github.com/Rdatatable/data.table/wiki).
By default, the verbs defined in this package automatically start a new frame whenever they want to define one of these elements,
but the current expression's frame has already specified it;
otherwise they add to the current frame.
More complex expressions are thus supported by automatically chaining `data.table` frames:

```{r chaining}
DT %>%
  start_expr %>%
  select(mean_mpg = mean(mpg)) %>%
  where(hp > 50L) %>%
  group_by(vs, am, gear) %>%
  order_by(gear, -vs, am) %>%
  filter(mean_mpg > 20) %T>%
  print %>%
  end_expr %>% {
    invisible(print(., nrows = 10L))
  }
```

If we wanted to be explicit about chaining whenever possible
(see [below](#explicit-chaining)),
we could set `options(table.express.chain = FALSE)`,
which would lead to a warning being shown whenever a part of the query is replaced.

### Verbs' effects in the frame

- Verbs that set `i`:
  - `arrange`
  - `filter`
  - `filter_on`
  - `filter_sd`
  - `order_by`
  - `where`
- Verbs that set `j`:
  - `mutate`
  - `mutate_sd`
  - `transmute`
  - `transmute_sd`
- Verbs for grouping:
  - `group_by` to specify `by`
  - `key_by` to specify `keyby`

## Explicit chaining

The automatic chaining mentioned above is not a problem in most situations.
For example the following chains lead to the same result,
and therefore have the same semantics:

```{r two-chains-one-semantic}
DT[mpg > 20, mpg * 2]

DT[mpg > 20][, mpg * 2]
```

However, these two chains have different semantics:

```{r two-chains-two-semantics}
DT[, .(mpg = mpg * 2)][mpg > 40]

DT[mpg > 40, .(mpg = mpg * 2)]
```

As mentioned [above](#expression-delimiters),
`chain` can be used to chain expressions by evaluating the current one with `end_expr`,
and immediately capturing the resulting `data.table` to start building a new expression.
This can be helpful in situations where automatic chaining
(or lack thereof)
can lead to a change in the expression's semantics:

```{r two-exprs-two-semantics}
DT %>%
  start_expr %>%
  transmute(mpg = mpg * 2) %>%
  filter(mpg > 40) %T>%
  print %>%
  end_expr

DT %>%
  start_expr %>%
  transmute(mpg = mpg * 2) %>%
  chain %>%
  filter(mpg > 40) %>%
  end_expr
```

### Eager verbs

In the following cases, the mentioned verbs use the captured `data.table` eagerly during expression building:

- `filter_sd` when `.SDcols` is a `tidyselect` call.
- `mutate_sd` when `.SDcols` is a `tidyselect` call.

This can lead to unexpected results if we don't keep in mind the expression that is built:

```{r implicit-chain-filter_sd}
DT[, mpg:disp]

DT %>%
  start_expr %>%
  select(mpg:disp) %>%
  transmute(ans = sqrt(mpg)) %>%
  filter_sd(.COL > 0, .SDcols = contains("m"))
```

The `select` gets rid of `am`,
but `filter_sd` sees the columns of `DT` before any expression has been evaluated.
Explicit chaining can help in these cases,
capturing intermediate results:

```{r explicit-chain-filter_sd}
DT %>%
  start_expr %>%
  select(mpg:disp) %>%
  chain %>%
  transmute(ans = sqrt(mpg)) %>%
  filter_sd(.COL > 0, .SDcols = contains("m"))
```
