---
title: "Using table.express"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using table.express}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library("table.express")
library("data.table")
library("magrittr")

# knitr defaults
knitr::opts_chunk$set(comment = "#>")

options("datatable.print.nrows" = 4L)
options("datatable.print.topn" = 2L)
```

The goal of this package is to offer an alternative way of **expressing** common operations with
[`data.table`](https://cran.r-project.org/package=data.table)
without sacrificing the performance optimizations that it offers.
The foundation for the data manipulation verbs is the
[`dplyr`](https://cran.r-project.org/package=dplyr)
package,
which also advocates the piping operator from the
[`magrittr`](https://cran.r-project.org/package=magrittr)
package.
The [`rlang`](http://cran.r-project.org/package=rlang)
package powers most of this package's functionality,
which means that
[tidy evaluation](https://tidyeval.tidyverse.org/)
can also be supported.
There are other resources that provide comprehensive descriptions of these packages,
so they will not be explained here.

Even though using data manipulation verbs can improve expressiveness in some cases,
this is not always true,
so using the traditional `data.table` syntax might still be preferable.

In order to ressemble SQL syntax more closely,
a couple of verb aliases are also defined:

- `where`, being equivalent to `dplyr`'s `filter`.
- `order_by`, being equivalent to `dplyr`'s `arrange`.

The examples here will be working with the `mtcars` data:

```{r mtcars}
data("mtcars")

DT <- mtcars %>%
    as.data.table %T>%
    print
```

## Expression delimiters

The foundation for this package is building expressions that are almost entirely delegated to `data.table`.
These expressions are built by *parsing* the input of the different verbs.
In order to explicitly show when an expression is being built and subsequently evaluated,
we use 3 delimiters:

- `start_expr`
- `end_expr`
- `chain`

These also serve as visual reminders that we are not dealing directly with `data.table`s during the process.
We capture the input `data.table` and start the process with `start_expr`,
and evaluate the final expression with `end_expr`.
Using `chain` is equivalent to calling `end_expr` immediately followed by `start_expr`,
and should be seldom needed.

## Arranging rows

The `arrange`/`order_by` verbs add an expression with `order` to the frame,
and let `data.table` handle it as usual:

```{r arrange}
DT %>%
    start_expr %>%
    order_by(mpg, -cyl) %T>%
    print %>%
    end_expr
```

We see here that the built expression includes a `.DT_` pronoun.
When the expression is evaluated,
the captured `data.table` is assigned to the evaluation environment as said pronoun.

## Selecting columns

Even though selecting a subset of columns is a common operation,
it may be undesirable to do so when working with `data.table`s because they lead to data copies.

```{r df-copy}
x <- 1:2
tracemem(x)

df <- data.frame(x=x)
x2 <- df[, "x"]
x2[1L] <- 0L
```

With this normal data frame,
only the last assignment triggered a copy.

```{r dt-copy}
dt <- data.table(x=x)
x3 <- dt[, x]
```

In this case with `data.table`,
more copies were triggered.
Given that `data.table` supports modification by reference,
these copies are necessary.

With that said,
the `select` verb can be used as usual:

```{r select}
DT %>%
    start_expr %>%
    select(mpg, am) %T>%
    print %>%
    end_expr
```

To maintain consistency,
even single columns are kept as `data.table`s:

```{r select-single}
DT %>%
    start_expr %>%
    select(mpg) %T>%
    print %>%
    end_expr
```

In the case of single expressions in `select`,
calls to `tidyselect`'s helpers or to `:` are handled specially internally:

```{r select-special}
DT %>%
    start_expr %>%
    select(mpg:cyl) %>%
    end_expr

DT %>%
    start_expr %>%
    select(contains("M", ignore.case = TRUE)) %>%
    end_expr
```

Tidy evaluation and the `.parse` argument can also aid in cases where certain parts of the query were computed programmatically:

```{r select-tidy}
selected <- c("mpg", "am")
DT %>%
    start_expr %>%
    select(!!!selected, .parse = TRUE) %>%
    end_expr
```

### Transmuting columns

Given the way `data.table` handles the `j` part of the "query",
creating and keeping only new columns
(like with `dplyr`'s `transmute`)
can be done with `select`,
so `transmute` is simply an alias during expression building.

```{r transmute}
DT %>%
    start_expr %>%
    select(foo = mpg * 2, bar = exp(cyl)) %>%
    end_expr

DT %>%
    start_expr %>%
    transmute(foo = mpg * 2, bar = exp(cyl)) %>%
    end_expr
```

## Mutating columns

The `mutate` verb builds an expression with `:=` in order to perform assignment by reference by default.
This can be avoided by passing `.by_ref = FALSE` to `end_expr`,
which will use `data.table::copy` before assigning `.DT_`:

```{r mutate}
DT %>%
    start_expr %>%
    mutate(mpg = mpg / 2, hp = log(hp))

DT %>%
    start_expr %>%
    mutate(mpg = mpg / 2, hp = log(hp)) %>%
    end_expr(.by_ref = FALSE) %>% {
        invisible(print(.))
    }

print(DT)
```

## Filtering rows

The `where`/`filter` verbs work with the `i` part of the query:

```{r filter}
DT %>%
    start_expr %>%
    filter(vs == 1L, carb > 2L) %T>%
    print %>%
    end_expr

DT %>%
    start_expr %>%
    select(mean_mpg = mean(mpg)) %>%
    where(vs == 1L, carb > 2L, .collapse = `|`) %T>%
    print %>%
    end_expr
```

The helper verb `filter_sd` can be used to apply the same conditions to many columns,
and it can use a special pronoun `.COL` while specifying the expression:

```{r filter-sd}
DT %>%
  start_expr %>%
  filter_sd(`>`, 20, .SDcols = c("mpg", "qsec")) %T>%
  print %>%
  end_expr

DT %>%
  start_expr %>%
  filter_sd(.COL > 20, .SDcols = c("mpg", "qsec")) %>%
  end_expr
```

## Modifying subset of data

In order to support functionality similar to `data.table`'s `DT[, lapply(.SD, fun), .SDcols = c("...")]` syntax,
2 `data.table`-specific verbs are provided: `mutate_sd` and `transmute_sd`.

Starting with `mutate_sd`,
one additional feature it provides is that columns in `.SDcols` are modified by reference,
and columns that are not part of `.SDcols` can still be kept:

```{r mutate-sd}
DT %>%
    start_expr %>%
    mutate_sd(exp, .SDcols = c("mpg", "cyl")) %>%
    end_expr

print(DT)
```

Moreover, `mutate_sd` supports the special `.COL` pronoun that symbolizes the column that should be modified,
and can be used to express the mutation expression:

```{r mutate-sd-col}
DT %>%
    start_expr %>%
    mutate_sd(log(.COL), .SDcols = c("mpg", "cyl")) %>%
    end_expr

print(DT)
```

On the other hand, `transmute_sd` never modifies by reference,
and supports special expressions to "build" what is chosen as `.SDcols`.
These expressions can use `tidyselect` helpers,
as well as another special pronoun: `.COLNAME`:

```{r transmute-sd}
DT %>%
    start_expr %>%
    transmute_sd(.COL * 2, .SDcols = starts_with("d")) %>%
    end_expr

DT %>%
    start_expr %>%
    transmute_sd(.COL * 2, .SDcols = grepl("^d", .COLNAME)) %>%
    end_expr
```

## Data manipulation by group

Since `data.table` already supports this by means of its `by` parameter,
the `group_by` verb simply parses its input and assigns it accordingly:

```{r group-by}
DT %>%
    start_expr %>%
    select(.N) %>%
    group_by(gear) %T>%
    print %>%
    end_expr
```

The `key_by` verb does the same but sets the key of the result in order to sort:

```{r key-by}
DT %>%
    start_expr %>%
    select(.N) %>%
    key_by(gear) %T>%
    print %>%
    end_expr
```

## Chaining expressions

Naturally, more complex expressions are supported by automatically chaining `data.table` frames:

```{r chaining}
DT %>%
    start_expr %>%
    select(mean_mpg = mean(mpg)) %>%
    where(hp > 50L) %>%
    group_by(vs, am, gear) %>%
    order_by(gear, -vs, am) %>%
    filter(mean_mpg > 20) %T>%
    print %>%
    end_expr %>% {
        invisible(print(., nrows = 10L))
    }
```
